Identifying Effective Motivational Interviewing Communication Sequences Using Automated Pattern Analysis

### 2018-09-18 jhir/latex-manuscript/manuscript_publish.tex
integrate April's comments into the final manuscript

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}. The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}. The relational hypothesis suggests that counselors' use of accurate empathy, positive regard and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for and/or commitment to behavior change. Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.

@@

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov Chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. However, in MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider \hl{the} similarities between behavior codes and, consequently, their inability to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. Therefore, there is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts and assessing intervention fidelity. Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework by observing counselor language during motivational interviews. Their system employed an SVM classifier based on n-grams (contiguous sequences of words of a specified length), syntactic and semantic features. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical, contextual and semantic features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}. The SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, which is comparable to human coders. In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. The current work builds on past work performed on automation of annotation. We examine the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining in identifying the counselor communication strategies antecedent to patient change talk.

@@

Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, which they developed to identify patterns in customer purchases. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}. However, no published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in a more compact form that are easier to interpret. In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is a state-of-the-art closed frequent itemset mining algorithm, which has demonstrated good performance in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis of pre-coded MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention. Namely, they limited their HMM model to three hidden states which they characterized as ``towards change'', ``away from change'' and ``non-determined''. These states were then used as predictors in a bivariate regression model to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov Chain-based methods most typically used in MI research. First-order Markov Chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

\subsection{Dataset}
\label{subsec:dataset}
The dataset for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had been previously annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} excluding conversations that correspond to greetings, farewell and interview setups such as table and camera settings. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: \hl{weight-related} statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from beginning to end was analyzed. Successful sequences were defined as those that resulted in a patient change talk or commitment language statement. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated using this approach. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which the initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of the difference between pattern occurrence in successful and unsuccessful communications was reported by computing \hl{the} p-value of Pearson's chi-square test.

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. We observed 4,724 state transitions in the successful sequences, where 4,679 and 45 state transitions occurred between different and same states, respectively. On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences. Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state most often transitioned to \textit{``active feedback''} (27\%) and back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, 1,106 state transitions occurred within the unsuccessful sequences, where 697 and 409 state transitions happened between different and same states, respectively. The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences. Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.

@@

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting \hl{the} success of motivational interviews at any point during the interview~\cite{hasan2018predicting}. In this paper, we present current work that builds on past work performed on automation of annotation, more specifically, the sequential analysis of patient-counselor communication data to identify the counselor communication strategies linked to the elicitation of change talk and sustain talk. Next steps include examining the performance of these models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.

>>>>

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}. The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}. The relational hypothesis suggests that counselors' use of accurate empathy, positive regard and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (``MI-consistent'' or MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for and/or commitment to behavior change~\cite{amrhein2004does}. Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.

@@

The sequential analysis procedure used in the above MI process studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} is based on the first-order Markov Chain model~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}. The Markov Chain model is a discrete-time stochastic process built on the assumption that the state of a system or condition changes over time and only depends on the previous event. Hence, Markov Chain models have two main drawbacks. The first is their inability to preserve the long-range dependencies between observations in a sequence. In MI, an observed behavior can be influenced by any of the preceding behaviors. The second drawback is their inability to consider similarities between behavior codes and, consequently, first-order Markov chain models are unable to identify multiple similar behaviors that lead to the same outcome. Thus, first-order Markov models may be insufficient to fully understand the associations between behaviors in patient-counselor communication sequences. There is a need for more powerful computational methods, which consider clusters of similar behavior codes and long-range dependencies between behaviors, to identify causal relationships. The goal of the current research is to test the applicability of data mining and machine learning methods to identify effective patterns of patient-counselor communication.

Several prior works have reported the results of adopting machine learning methods, such as topic models~\cite{kotov2014towards, gaut2017content, imel2015computational, kotov2015interpretable, atkins2014scaling}, classification methods~\cite{hasan2016study, can2016sounds, perez2017predicting, perez2017understanding} and neural networks~\cite{hasan2016study, tanana2015recursive, tanana2016comparison} to the tasks of annotating MI transcripts for the assessment of intervention fidelity. Perez-Rosas et al.~\cite{perez2017predicting} developed a natural language processing system to evaluate counselor fidelity to the MI framework. Their system employed a Support Vector Machines (SVM) classifier based on n-grams (contiguous sequences of words of a specified length), syntactic (structure of the clinician statements) and semantic (cognitive state) features. In our own recent work, we evaluated the accuracy of state-of-the-art classification methods and deep neural networks in conjunction with the lexical (words expressed), contextual (prior code) and semantic (inferred cognitive state based on Linguistic Inquiry and Word Count dictionaries~\cite{tausczik2010psychological}) features for the task of automated annotation of MI transcripts using codebooks with varying numbers of behavior codes~\cite{hasan2016study}. An SVM model with the aforementioned features achieved 75\% accuracy for automated annotation of MI transcripts with 17 behavior codes, accuracy comparable to human coders. In a follow-up study~\cite{hasan2018predicting}, we applied Markov Models (MMs) and Recurrent Neural Networks (RNNs) to the classification of coded patient-counselor communication sequences into successful (sequences leading to patient change talk) and unsuccessful (sequences resulting in sustain talk). RNN achieved 87\% accuracy, 17\% greater than MMs (70\%) in predicting the success of motivational interviews. The current work builds on our past work by examining the efficacy of Hidden Markov Models (HMMs) and frequent pattern mining for the identification of the counselor communication strategies leading to patient change talk.

@@

Frequent pattern mining~\cite{aggarwal2014frequent} is a class of data mining methods to identify sets of items (or observations, referred to as itemsets) which frequently appear together. Agrawal and Srikant~\cite{agrawal1994fast} first introduced frequent pattern mining with the Apriori algorithm, developed to identify customer purchasing patterns. Since its introduction, frequent pattern mining has been applied to several other domains, including health informatics~\cite{abdullah2008analysis, olukunle2002fast, bethel2006mining, wright2015use}, medical imaging~\cite{olukunle2002fast}, chemical and biological analysis~\cite{deshpande2005frequent, leach2007introduction, zhang2008discovering}, web mining~\cite{srivastava2000web} and outlier analysis~\cite{aggarwal2015outlier}. However, no published research has yet examined the utility of frequent pattern mining for studying patient-counselor communication. A major challenge in applying frequent pattern mining methods to patient-counselor communication sequences is the large number of resulting patterns, which include redundant patterns. To address this problem, we utilized the closed frequent itemset mining method~\cite{pasquier1999discovering}, which produces fewer patterns in a more compact form that are easier to interpret. In this study, we leveraged FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining method, to identify the counselor behaviors that frequently lead to patient change talk. FPClose is a state-of-the-art closed frequent itemset mining algorithm, which has demonstrated good performance in terms of running time and memory consumption.

In this paper, we focus on computational methods to facilitate the sequential analysis of pre-coded MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns will provide empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper is the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM to identify hidden states in a brief motivational intervention. Limiting their HMM model to three hidden states which were characterized as ``towards change'', ``away from change'' and ``non-determined'', these states were used to predict drinking outcomes 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study was to evaluate the utility of using HMM and frequent pattern mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov Chain-based methods most typically used in MI research. First-order Markov Chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

\subsection{Dataset}
\label{subsec:dataset}
The dataset for this study consists of 37 audio-recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had been previously annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} excluding conversations that correspond to greetings, farewell and interview setups such as table and camera settings. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: weight-related statements about actions of commitment, change talk and ambivalence that occurred in the past, patient questions to the counselor and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
HUPO & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
LUP & Low uptake  & An utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

\subsection{Data preprocessing}
\label{subsec:datapreprocessing}
Utterances in MI session transcripts were segmented into successful and unsuccessful communication sequences. For each MI transcript, the stream of behavior codes from the beginning of a session to the end of the session was analyzed. Successful sequences were defined as those that resulted in a patient change talk or commitment language statement. Unsuccessful sequences were similarly created for sequences resulting in sustain talk. A total of 1,360 sequences were generated using this approach. The majority of the sequences (n=1,102) were successful, which is expected for a treatment-seeking population, in which patients initial motivation for behavior change is typically high. Successful sequences had an average length of 5.28 utterances, while unsuccessful sequences had on average 5.29 utterances.

@@

\subsubsection{Frequent Pattern Mining}
\label{subsubsec:frequentPatternMining}
We applied frequent pattern mining to identify frequently occurring patterns of patient-counselor behavior codes in successful and unsuccessful communication sequences. Behavior codes in these patterns may be separated by one or more other codes. For this purpose, we utilized FPClose~\cite{grahne2005fast}, an efficient state-of-the-art closed frequent pattern mining algorithm implemented in SPMF~\cite{fournier2014spmf, fournier2016spmf}, to identify frequent patterns of patient-counselor communication behaviors in successful and unsuccessful MI communication sequences. SPMF is an open-source library providing more than 150 data mining algorithms. Popular non-closed frequent pattern mining algorithms include Apriori~\cite{agrawal1994fast} and FP-Growth~\cite{han2000mining}. A \textit{frequent pattern} is defined as a pattern of observations, which appears in a given set of sequences more often than a user-specified threshold called the \textit{minimum support count}. For example, \{A\}, \{C\}, \{D\} and \{C, D\} are frequent patterns in the example set of sequences in Figure~\ref{fig:fpm}, since these patterns appear at least 2 times, which is the minimum support count in this example. In this work, we identified and analyzed \textit{closed frequent patterns} among all sequences of behavior codes in successful and unsuccessful communication sequences. A \textit{frequent pattern} is \textit{closed} if none of its supersets have the same support count~\cite{pasquier1999discovering}, where a set X is a superset of another set Y, if X contains all the elements of the set Y. For example, the itemsets \{A\}, \{C\}, \{D\} and \{C, D\} in Figure~\ref{fig:fpm} are not closed frequent patterns since their supersets \{A, B\}, \{B, C\}, \{B, D\} and \{B, C, D\} have the same support count. Therefore, \{B\}, \{A, B\}, \{A, D\}, \{B, C\}, \{B, D\} and \{B, C, D\} are closed frequent patterns since none of their supersets have the same support count. On the other hand, itemsets \{A, C\}, \{A, B, C\}, \{A, B, D\}, \{A, C, D\} and \{A, B, C, D\} have support counts of 1, 0, 1, 0 and 1, respectively, which is less than the minimum support count and thus are identified as non-frequent itemsets. Since the threshold for minimum support count depends on a task and is typically determined by the domain expert, we followed prior work~\cite{moradi2017quantifying, liu1999mining} and set the minimum support count as 10\% of the total number of all communication sequences, which is 110 for successful and 25 for unsuccessful communication sequences. For each pattern, the statistical significance of the difference between successful and unsuccessful sequences was computed with Pearson's chi-square test.

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of the HMM models are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. Three behaviors represented 45-69\% of each state's emission probability mass and, thus, were used to interpret the emission matrix and label the hidden states. We observed 4,724 state transitions in the successful sequences; 4,679 and 45 state transitions occurred between different and same states, respectively. On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences. Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state most often transitioned to \textit{``active feedback''} (27\%) and back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

In contrast, 1,106 state transitions occurred within the unsuccessful sequences; 697 and 409 state transitions happened between different and same states, respectively. The majority of unsuccessful sequences (98\%) began in a state of \textit{``ambivalence''} as indicated by the greater proportion of counselor reflections of both change talk (20\%) and sustain talk (12\%) as well as affirmations (13\%). About 2\% of the time unsuccessful sequences started in a state of \textit{``avoidance''}. Higher rates of patient low uptake (29\%) and other-related high uptake (11\%) statements and counselor patient-centered information (18\%) distinguished \textit{``avoidant''} sequences. Both \textit{``ambivalent''} (84\%) and \textit{``avoidant''} (61\%) states most frequently transitioned to the \textit{``avoidant''} state.

@@

This study is part of a line of research to develop machine-learning models to annotate (code) and analyze patient-counselor communication patterns. We have previously reported on the development of probabilistic generative models~\cite{kotov2014towards, kotov2015interpretable} and application of novel features for maximum margin and deep learning classifiers~\cite{hasan2016study} with the goal of automated annotation of MI session transcripts. Experiments applying the annotation model to novel datasets are underway to assess the generalizability of the model to more diverse types of clinical encounters (e.g, email coaching to increase fruit and vegetable intake, HIV clinical care visits~\cite{carcone2018using}). We also developed and evaluated probabilistic and deep learning methods for the task of predicting the change talk at any point during the motivational interview~\cite{hasan2018predicting}. Our current work builds on this past work to automatically annotate clinical encounters, specifically, the current study presents two approaches for the sequential analysis of patient-counselor communication data for the purpose of identifying the counselor communication strategies linked to the elicitation of change talk and sustain talk. Next steps include examining the performance of the HMM and frequent pattern mining models in diverse data sets representing different populations and behavioral problems. We are also experimenting with machine learning methods to segment (parse) the stream of communications in a healthy eating promotion intervention delivered by email into semantically coherent chunks (phrases, sentences or paragraphs), to which the annotation model can then assign a behavior code. Together, these models form the basis of a complete system to automatically code and analyze patient-counselor interactions. An automated system for behavioral coding and analysis could substantially accelerate the pace of research on the causal mechanisms of Motivational Interviewing and inform both the theory and clinical practice by providing clinicians with information about how to best tailor their communication strategies to different patient populations.
